---
title: "Unzipper"
author: "Mike Proctor"
date: "`r Sys.Date()`"
format: gfm


---

This script, is designed to efficiently unzip  geospatial archives (ZIP files) 
from the defined source_path. It performs a selective extraction based on the internal 
folder structure of each ZIP: all Shapefiles (SHP) are flattened and aggregated 
into a single unzipped/shp folder, while all File Geodatabases (FGDB) are 
extracted, and their unique .gdb folder structure is preserved and moved into 
the unzipped/fgdb folder, preventing data overwrites during the large-scale, 
concurrent extraction.

Gemini was abused extensively in the composition of these scripts.

There is another version that uses parallel processing and is much faster.

These were used primarily in working with files of land parcels downloaded from [TxGIO]<https://data.geographic.texas.gov/> using their bulk downloader.  The 
zipfiles had shp files of parcels for each county as well as gdb files for each 
county and one for the entire state.  The entire state is a little large to 
work with conviently so I removed it before unzipping.  The destination folder 
structure is the complex part of the script.  The gdb files for each county have
to retain their name and get put in a fgdb folder.  The shape files were easier 
to work with - they just ended up in a shp folder. 



```{r}



library(rprojroot)
library(tictoc)
library(beepr)

base_path       <- find_rstudio_root_file()
source_path     <- file.path(base_path, "source_data//")
unzip_path        <- file.path(base_path, "unzipped//")
zip_path        <- file.path(base_path, "zipped//")


# Define the final destination folders for the extracted contents
shp_destination_path <- file.path(unzip_path, "shp")
fgdb_destination_path <- file.path(unzip_path, "fgdb")

# Create the necessary destination folders
dir.create(source_path, recursive = TRUE, showWarnings = FALSE)
dir.create(shp_destination_path, recursive = TRUE, showWarnings = FALSE)
dir.create(fgdb_destination_path, recursive = TRUE, showWarnings = FALSE)

cat("--- Setup Complete ---\n")
cat("Zips are expected in: ", zip_path, "\n")
cat("Shapefiles will be extracted to: ", shp_destination_path, "\n")
cat("Geodatabases will be extracted to: ", fgdb_destination_path, "\n\n")



# Get a list of all .zip files in the source directory
zip_files <- list.files(
  path = zip_path,
  pattern = "\\.zip$", # Ensure only zip files are selected
  full.names = TRUE # Get the full path for the unzip function
)

if (length(zip_files) == 0) {
  stop("No .zip files found in the source_path. Please check the path and file existence.")
}

cat("Found", length(zip_files), "zip files. Starting selective extraction...\n")

```

```{r}

# --- 3. Loop and Perform Selective Unzip ---
tic()
sapply(zip_files, function(zip_file) {
  
  zip_name <- basename(zip_file)
  cat("\nProcessing:", zip_name, "\n")
  
  # A. Get the list of all files inside the current zip archive
  internal_files <- unzip(zip_file, list = TRUE)$Name
  
  # B. Extract FGDB files (Preserve folder structure!)
  fgdb_files_to_extract <- grep("^fgdb/", internal_files, value = TRUE)
  
  if (length(fgdb_files_to_extract) > 0) {
    # 1. Temporarily extract the GDB to a scratch folder
    temp_extract_path <- file.path(unzip_path, "temp_gdb_extract")
    dir.create(temp_extract_path, showWarnings = FALSE)
    
    unzip(
      zipfile = zip_file,
      files = fgdb_files_to_extract,
      exdir = temp_extract_path # Extract to a temporary folder
    
    )
    
    # 2. Identify the unique *.gdb folder that was just extracted (e.g., in temp_extract_path/fgdb)
    extracted_fgdb_root <- file.path(temp_extract_path, "fgdb")
    gdb_folder_name <- list.files(extracted_fgdb_root, pattern = "\\.gdb$", full.names = FALSE)
    
    if (length(gdb_folder_name) == 1) {
      # 3. Move the unique GDB folder to the final destination
      source_gdb_path <- file.path(extracted_fgdb_root, gdb_folder_name)
      destination_gdb_path <- file.path(fgdb_destination_path, gdb_folder_name)
      
      file.rename(from = source_gdb_path, to = destination_gdb_path)
      cat(sprintf("  -> Extracted and moved GDB: %s\n", gdb_folder_name))
      
    } else {
       cat("  -> Warning: Could not find unique *.gdb folder to move.\n")
    }
    
    # 4. Clean up the temporary directory
    unlink(temp_extract_path, recursive = TRUE)
    
  } else {
    cat("  -> No 'fgdb' folder/files found for extraction.\n")
  }

  # C. Extract SHP files (Flatten structure!)
  shp_files_to_extract <- grep("^shp/", internal_files, value = TRUE)
  
  if (length(shp_files_to_extract) > 0) {
    unzip(
      zipfile = zip_file,
      files = shp_files_to_extract, 
      exdir = shp_destination_path, 
      junkpaths = TRUE # Use junkpaths=TRUE to flatten 'shp/' contents
    )
    cat(sprintf("  -> Extracted %d SHP files to %s\n", length(shp_files_to_extract), basename(shp_destination_path)))
  } else {
    cat("  -> No 'shp' folder/files found for extraction.\n")
  }
  
  return(TRUE)
})

toc()
beep(sound = 4)

# 235 files - 145.28 sec

```

